---
title: "Lab 04 - Data Visualization"
output:
  html_document:
    html_preview: false
link-citations: yes
---

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(leaflet)
library(tidyverse)
```

# Learning Goals

- Read in and prepare the meteorological dataset
- Create several graphs with different `geoms()` in `ggplot2`
- Create a facet graph
- Conduct some customizations of the graphs
- Create a more detailed map using `leaflet()`


# Lab Description

We will again work with the meteorological data presented in lecture.

**The objective of the lab is to examine the association between weekly average dew point temperature and wind speed in four regions of the US and by elevation.**

# Steps

### 1. Read in the data

First download and then read in with `data.table:fread()`

```{r checking-file}
fn <- paste0("https://raw.githubusercontent.com/JSC370/",
             "jsc370-2023/main/labs/lab03/met_all.gz")
if (!file.exists("met_all.gz")) # avoid downloading again
  download.file(fn, destfile = "met_all.gz")
met <- data.table::fread("met_all.gz")
```

### 2. Prepare the data

- Remove temperatures less than -17C
- Make sure there are no missing data in the key variables coded as 9999, 999, etc
- Generate a date variable using the functions `as.Date()` (hint: You will need the following to create a date `paste(year, month, day, sep = "-")`).
- Using the `data.table::week` function, keep the observations of the first week of the month.
- Compute the mean by station of the variables `temp`, `rh`, `wind.sp`, `vis.dist`, `dew.point`, `lat`,`lon`, and `elev`.
- Create a categorical variable for elevation (break point at 252 m)
- Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees

```{r}
# remove temperatures less than or equal to -17C / code 9999 elev as NA
met <- met[temp > -17][elev == 9999, elev := NA]
# create the week variable
met[ , week := week(as.Date(paste(year, month, day, sep = "-")))]
# select the first week
met_first_week <- met[week == min(week, na.rm = TRUE)]
# group and compute means
met_avg <- met_first_week[ , .(
  temp = mean(temp, na.rm = TRUE),
  rh = mean(rh, na.rm = TRUE),
  wind.sp = mean(wind.sp, na.rm = TRUE),
  vis.dist = mean(vis.dist, na.rm = TRUE),
  dew.point = mean(dew.point, na.rm = TRUE),
  lat = mean(lat), lon = mean(lon),
  elev = mean(elev, na.rm = TRUE)
), by = "USAFID"]
# categorize elevation
met_avg[ , elev_cat := ifelse(elev > 252, "high", "low")]
# categorize into 4 regions
met_avg[ , region := ifelse(
  lon > -98 & lat > 39.71, "North East", ifelse(
    lon > -98, "South East", ifelse(
      lat > 39.71, "North West", "North East"
    )
  )
)]
# simpler (cred: Kevin Wang)
met_avg[ , region := paste(
  ifelse(lat > 39.71, "North", "South"),
  ifelse(lon > -98, "East", "West")
)]
# check counts
table(met_avg$region, useNA = "always")
```

```{r}
# tidyverse version
met_avg_tidy <- met |>
  filter(temp > -17) |>  # keep temperater > -17C
  mutate(
    elev = ifelse(elev == 9999, NA, elev),  # use NA to mark missing data
    week = week(as.Date(paste(year, month, day, sep = "-"))) # create week 
  ) |>
  filter(week == min(week, na.rm = TRUE)) |> # select first week records only
  group_by(USAFID) |>  # summarise per station
  summarise(across(
    c(temp, rh, wind.sp, vis.dist, dew.point, lat, lon, elev),
    ~ mean(.x ,na.rm = TRUE)
    )) |>  
  mutate( # categorise elevation and region
    elev_cat = if_else(elev > 252, "high", "low"),
    region = case_when(
      lon > -98 & lat > 39.71 ~ "North East",
      lon > -98  ~ "South East",
      lat > 39.71 ~ "North West",
      TRUE ~ "South West"
    )
  )
table(met_avg_tidy$region, useNA = "always")
```

### 3. Use `geom_violin` to examine the wind speed and dew point temperature by region

You saw how to use `geom_boxplot` in class. Try using `geom_violin` instead (take a look at the help).
(hint: You will need to set the `x` aesthetic to 1)

- Use facets
- Make sure to deal with `NA` category
- Describe what you observe in the graph

```{r}
met_avg |>
  filter(!is.na(region), !is.na(wind.sp)) |>
  ggplot() +
  geom_violin(mapping = aes(y = wind.sp, x = 1)) +
  facet_wrap(~ region, nrow = 2)
```

```{r}
met_avg |>
  filter(!(region %in% NA), !is.na(dew.point)) |>
  ggplot() +
  geom_violin(mapping = aes(y = dew.point, x = 1, fill = region)) +
  facet_wrap(~region, nrow = 2)
```


### 4. Use `geom_jitter` with `stat_smooth` to examine the association between dew point temperature and wind speed by region

- Colour points by region
- Make sure to deal with `NA` category
- Fit a linear regression line by region
- Describe what you observe in the graph

```{r}
met_avg |>
  filter(!is.na(region), !is.na(dew.point), !is.na(wind.sp)) |>
  ggplot(mapping = aes(x = wind.sp, y = dew.point, color = region))+
  stat_smooth(method = lm, formula = y ~ x) +
  geom_jitter() 
```

_Tips on reducing visual clutter_

+  Try using facets. 
+  Remove the intervals for the smoothing lines using `se = FALSE`. 
+  Make the points semi-transparent setting `alpha` to a number between 0 and 1 to further reveal overlapped points.

```{r}
met_avg |>
  filter(!is.na(region), !is.na(dew.point), !is.na(wind.sp)) |>
  ggplot(mapping = aes(
    x = wind.sp, y = dew.point, color = region))+
  geom_jitter(alpha = .2) +
  stat_smooth(method = lm, formula = y ~ x, se = FALSE) +
  facet_wrap(~ region)
```


### 5. Use `geom_bar` to create barplots of the weather stations by elevation category coloured by region

- Bars by elevation category using `position = "dodge"`
- Change colors from the default. Color by region using `scale_fill_brewer` see [this](http://rstudio-pubs-static.s3.amazonaws.com/5312_98fc1aba2d5740dd849a5ab797cc2c8d.html)
- Create nice labels on axes and add a title
- Try a different theme
- Describe what you observe in the graph
- Make sure to deal with `NA`

```{r}
met_avg |>
  filter(!is.na(region), !is.na(elev_cat)) |>
  ggplot() +
  # theme_minimal() +
  # theme_bw() +
  # theme_classic() +
  # you can remove and customize individual components of the plot
  theme_void() + theme(axis.text = element_text()) +
  geom_bar(mapping = aes(x = elev_cat, fill = region), 
           position = "dodge")+
  # scale_fill_viridis_d()
  # scale_fill_brewer(palette = "PuOr")
  scale_fill_brewer(palette = "Paired") +
  scale_x_discrete(labels = c("High\n(>252m)", "Low\n(<252m)")) +
  labs(
    title = "Number of weather stations by elevation category and region",
    x = "Elevation", y = "Count", fill = "Region"
    )
```

### 6. Use `stat_summary` to examine mean dew point and wind speed by region with standard deviation error bars

- Make sure to remove `NA`
- Use `fun.data="mean_sdl"` in `stat_summary`
- Add another layer of `stats_summary` but change the geom to `"errorbar"` (see the help).
- Describe the graph and what you observe

```{r}
met_avg |>
  filter(!is.na(region), !is.na(dew.point)) |>
  ggplot(mapping=aes(x = region, y = dew.point)) +
  stat_summary(fun.data = "mean_sdl") +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar") +
  labs(x = "", y = expression("Dew Point Temperature ("*degree*C*")"),
       title = "Mean dew point temperature and 2SD error bar per region") +
  theme_minimal()
```

_`expression()` allows mathemmatical expressions in plots. See [here](https://r-graphics.org/recipe-annotate-text-math) for other methods in ggplot_.

```{r}
met_avg |>
  filter(!is.na(region), !is.na(wind.sp)) |>
  ggplot(mapping = aes(x = region, y = wind.sp)) +
  stat_summary(fun.data = "mean_sdl") +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar") +
  labs(x = "", y = "Wind Speed (mph)",
       title = "Mean wind speed and 2SD error bar per region") +
  theme_minimal()
```

- Dew point temperature is...
- Wind speed is...

### 7. Make a map showing the spatial trend in relative humidity (`rh`) in the US

- Make sure to remove `NA`
- Use `leaflet()`
- Make a colour palette with custom colours
- Use `addMarkers` to include the top 10 places in relative h (hint: this will be useful `rank(-rh) <= 10`)
- Add a legend
- Describe trend in RH across the US

```{r}
# prepare data
met_avg2 <- met_avg[!is.na(rh)]
top10 <- met_avg2[rank(-rh) <= 10]
```

```{r}
rh_pal <- colorNumeric(
  # c('lightgrey','lightblue','blue'),
  viridisLite::viridis(5, direction = -1),
  domain = met_avg2$rh
  )
leaflet(met_avg2) %>%
  addProviderTiles('OpenStreetMap') |>
  addCircles(lat = ~lat, lng = ~lon, color = ~rh_pal(rh),
             label = ~paste0(round(rh,2), ' rh'), 
             opacity = 1, fillOpacity = 1, radius = 250) |>
  addMarkers(lat = ~lat, lng = ~lon, 
             label = ~paste0(round(rh, 2), ' rh'), 
             data = top10) |>
  addLegend('bottomleft', pal = rh_pal, values = met_avg2$rh, 
            title = "Relative Humidity (%)", opacity=1)
```


### 8. Use a ggplot extension

- Pick and extension (except `cowplot`) from [here](https://exts.ggplot2.tidyverse.org/gallery/) and make a plot of your choice using the `met` data (or `met_avg`)
- Might want to try examples that come with the extension first (e.g. `ggtech`, `gganimate`, `ggforce`)

**`ggridges` example**

+ `fig.asp` sets the figure aspect ration in the output.

```{r fig.asp=2}
library(ggridges)
met |>
  group_by(USAFID, day) |>
  summarise(temp = mean(temp, na.rm = TRUE), .groups = "drop") |>
  ggplot() +
  theme_minimal() +
  geom_density_ridges(
    aes(x = temp, y = day, group = day),
    alpha = .5, bandwidth = .3
    ) +
  scale_y_reverse(breaks = c(1, 15, 31)) +
  labs(x = expression("Temperature ("*degree*C*")"), y = "", 
       title = "Daily temperature distributions in August, 2019 across the US")
```

**`gganimate` + `ggbeeswarm` example**

```{r}
library(gganimate)
library(ggbeeswarm)
met_daily_avg <- met |>
  mutate(
    region = paste(
      ifelse(lat > 39.71, "North", "South"),
      ifelse(lon > -98, "East", "West"))
  ) |>
  group_by(region, day, USAFID) |>
  summarise(temp = mean(temp, na.rm = TRUE), .groups = "drop") 
animated_met <- met_daily_avg |>
  ggplot() +
  theme_void() + 
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 14)
    ) +
  geom_beeswarm(aes(y = temp, x = region, colour = temp),
                cex = .75, size = .5, show.legend = FALSE) +
  scale_colour_viridis_c() +
  # animate
  labs(y = "", x = expression("Temperature ("*degree*C*")"),
       title = "Day: {frame_time}") +
  transition_time(day) +
  ease_aes("circular-in-out")
animate(animated_met, fps = 100, nframes = 3100 / 5)
```
