---
title: "Advanced Regression"
author: "Meredith Franklin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mgcv)
library(ggplot2)
co2<-read.csv("co2_mm_mlo.csv")
```

## Advanced Regression

- A linear model tries to fit the best straight line that passes through the data, so it doesn't work well for all datasets.

- In contrast, a GAM can capture complex relationships by fitting a non-linear smooth function through the data, while controlling how wiggly the smooth can get

- Let's use the CO2 dataset that has been collected at the top of Mauna Loa, Hawaii since 1958. The data were dowloaded from [here](https://gml.noaa.gov/ccgg/trends/). This 

```{r}
head(co2)
```

- Let's add a linear and a smooth

```{r}
ggplot(co2, aes(x=dec_date, y=co2))+
    geom_point()+
    geom_smooth(method='lm',col="red")+
    geom_smooth(col="blue")
```
- Subset the data to see seasonal trends

```{r}
co2_s<-co2[co2$year >= 2010,]
ggplot(co2_s, aes(x=dec_date, y=co2))+
    geom_point()+
    geom_smooth(method='lm',col="red")+
    geom_smooth(col="blue")
```
- Look at the linear model fit of the subset data
```{r}

```
 - Use the linear model to predict CO2
 
 ```{r}

```

 - Look at the gam model on the subset data and predict CO2

```{r}

```
- We see that it is still not doing much. We need to capture the within year cycles in the data, so let's add a smooth on month

- That is, fit to all data and smoothing date (overall trends) and month (to get within year trends)

```{r}


```

